Large squeezing in multicomponent trapped Bose-Einstein condensates

### 2013-06-11 large_number_squeezing_with_atoms.tex
Going back and cleaning up the Bog section.

We can understand the origin and scaling of this degradation of squeezing by considering the response of the condensate to small fluctuations about the mean field.  We do this using Bogoliubov theory \cite{PethickSmith}, which considers the first order quantum-mechanical fluctuations about the mean field.  To simplify the analysis we consider the case in which $U_{aa} \neq 0$, $U_{bb}=U_{ab} = 0$ and the condensate has a spatially-constant density in a box of side lengths $L_x$, $L_y$ and $L_z$.  In this limit the occupation of the non-zero momentum modes in the $\ket{a}$ internal state at the end of the pulse sequence is (a derivation is given in \ref{appendixBogDerivation})
\begin{align}
  n_a(\mathbf{k}, t_3) &= \expect{\hat{a}^\dagger(\mathbf{k},t) \hat{a}(\mathbf{k},t)} = \left[n_a \chi_{aa} \tau_\text{hold} \cos(\theta) \sinc\left(\omega_\mathbf{k} \tau_\text{hold}\right) \right]^2, \label{eqBogOccupation}
\end{align}
where the Bogoliubov mode frequency is $\omega_\mathbf{k} = \sqrt{\omega^0_\mathbf{k}(\omega^0_\mathbf{k} + 2 \chi_{aa} n_a)}$, and the free particle frequency is $\omega^0_\mathbf{k} = \hbar \mathbf{k}^2/2 M$.  The Bogoliubov modes can be expected to have an impact on the squeezing of the system when their occupation is a non-negligible fraction of the total number of atoms.  The proportional occupation of the Bogoliubov modes (i.e.\ the condensate depletion) is
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &= \frac{1}{N}\sum_{\mathbf{k}\neq 0} n_a(\mathbf{k}, t), \label{eqCondensateDepletion}
\end{align}
where the sum over $\mathbf{k} \neq 0$ is taken over the available non-zero momentum modes $\mathbf{k} = (2\pi/L_x) n_x \hat{\mathbf{x}} + (2\pi/L_y) n_y \hat{\mathbf{y}} + (2\pi/L_z) n_z \hat{\mathbf{z}}$ with $n_x$, $n_y$ and $n_z$ arbitrary integers (not all zero).

[Insert motivation that the impact of Bogoliubov modes is detrimental here.  I could wave my hands and say that more modes is bad, as seen in the previous parts of the article, or I could refer to the squeezing of the Bogoliubov modes:
\begin{align}
  S_{a,\mathbf{k}\neq 0}(t_3) &= \frac{\frac{1}{2}\sum_{\mathbf{k}\neq0}n_a(\mathbf{k},t_3)\left[3 + \cos 2\theta + n_a(\mathbf{k}, t_3)\right]}{\sum_{\mathbf{k}\neq0} n_a(\mathbf{k}, t_3)},
\end{align}
which is always greater than 1, but this isn't convincing either as the total squeezing includes the number variance of the $\mathbf{k=0}$ mode as well as the correlations between the $\mathbf{k}=0$ mode and the $\mathbf{k}\neq 0$ modes.  If you do it naively, you find that when the Bog modes have a significant impact on the squeezing, they have also significantly changed it at $t=t_2$, which must be the same as $t=t_1$.  The problem is that the Bogoliubov theory used isn't number-conserving.  The best argument I have at the moment for Bog theory being a good indicator for when things go bad is that a significant Bog fraction implies a depleted condensate, and that will almost certainly make squeezing bad as it will invalidate the single-mode theory.  More freedom == bad?  In this case, the best evidence for the utility of Bog theory would seem to be Figure~\ref{fig1D_squeezing_and_bog_scaling}.]

Good squeezing requires that the Bogoliubov fraction or condensate depletion be small.  To better understand the scaling of $N_{\mathbf{k}\neq0}/N$, we will approximate the sum in \eqref{eqCondensateDepletion} with an integral.  In 1D \eqref{eqCondensateDepletion} becomes
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &= \frac{(n_a \chi_{aa} \tau_\text{hold} \cos\theta)^2}{N} \sum_{k \neq 0} \sinc^2\left(\sqrt{\omega^0_{k}(\omega^0_{k}+2 \chi_{aa}n_a)}\tau_\text{hold}\right).
\end{align}
In the limit that $\omega_{k_\text{min}} \tau_\text{hold} \ll 1$, where $k_\text{min} = 2\pi/L$ this discrete sum may be approximated by the integral
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{(n_a \chi_{aa} \tau_\text{hold} \cos\theta)^2}{N} \int_{-\infty}^{+\infty} \sinc^2 \left(\sqrt{\omega^0_{k}(\omega^0_{k}+2 \chi_{aa}n_a)}\tau_\text{hold}\right) \frac{d k}{\Delta k},
\end{align}
where $\Delta k = 2\pi/L$.  Changing integration variables to $\Phi = \omega^{0}_{\mathbf{k}}\tau_\text{hold}$, the integral can be written as
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{(n_a \chi_{aa} \tau_\text{hold} \cos\theta)^2}{N} \frac{L}{\sqrt{2}\pi\sqrt{\frac{\hbar}{M}\tau_\text{hold}}} \int_0^{\infty} \frac{1}{\sqrt{\Phi}}\sinc^2\sqrt{\Phi(\Phi + 2 \lambda_{aa} n_a)}\, d\Phi
\end{align}
Define $\Lambda_{a} = n_a \chi_{aa} \tau_\text{hold}$ to simplify the expression and we get
\begin{align}
  \frac{N^\text{1D}_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{1}{N} \frac{\Lambda_a^2 L \cos^2\theta } {\sqrt{2}\pi\sqrt{\frac{\hbar}{M}\tau_\text{hold}}} f^\text{1D}(\Lambda_a),
\end{align}
where
\begin{align}
  f^\text{1D}(\Lambda) &= \int_0^{\infty} \frac{1}{\sqrt{\Phi}}\sinc^2\sqrt{\Phi(\Phi + 2 \Lambda)}\, d\Phi
\end{align}

In 2D and 3D you get
\begin{align}
  \frac{N^{\text{2D}}_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{1}{N} \frac{\Lambda_a^2 A \cos^2\theta } {4\frac{\hbar}{M}\tau_\text{hold}} f^\text{2D}(\Lambda_a), \\
  \frac{N^{\text{3D}}_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{1}{N} \frac{\Lambda_a^2 V \cos^2\theta } {\sqrt{2}\pi^2\left(\frac{\hbar}{M}\tau_\text{hold}\right)^{3/2}} f^\text{3D}(\Lambda_a),
\end{align}
where
\begin{align}
  f^\text{2D}(\Lambda) &= \int_0^{\infty} \sinc^2\sqrt{\Phi(\Phi + 2 \Lambda)}\, d\Phi\\
  f^\text{3D}(\Lambda) &= \int_0^{\infty} \sqrt{\Phi} \sinc^2\sqrt{\Phi(\Phi + 2 \Lambda)}\, d\Phi
\end{align}

@@

@@

\begin{figure}
  \centering
  \begin{subfigure}{.45\textwidth}
    \centering
    Plot of the number occupation of the condensate and non-zero momentum modes as computed by the Bog theory and a simulation.  The simulation agrees with the Bog theory for the occupation of the non-zero momentum modes, but there is significant ~10-20\% depletion of the condensate.  The horizontal axis is the hold time.
  \end{subfigure}
  \begin{subfigure}{.45\textwidth}
    \centering
Squeezing curve showing the two-mode model, Bogoliubov squeezing \eqref{eqBogoliubovSqueezing} and simulation.  All three disagree, but both the Bog squeezing and simulation are worse than the two-mode model.  The same parameters are used as in the first part of this figure
  \end{subfigure}%
\caption{This is not a caption.}
  \label{figBogSqueezingValidation}
\end{figure}

>>>>

We can understand the origin and scaling of this degradation of squeezing by considering the response of the condensate to small fluctuations about the mean field.  We do this using Bogoliubov theory \cite{PethickSmith}, which considers the first order quantum-mechanical fluctuations about the mean field.  To simplify the analysis we consider the case in which $U_{aa} \neq 0$, $U_{bb}=U_{ab} = 0$ and the condensate has a spatially-constant density in a box of side lengths $L_x$, $L_y$ and $L_z$.  In this limit the occupation of the non-zero momentum modes in the $\ket{a}$ internal state at the end of the pulse sequence is (a derivation is given in \ref{appendixBogDerivation})
\begin{align}
  n_a(\mathbf{k}, t_3) &= \expect{\hat{a}^\dagger(\mathbf{k},t) \hat{a}(\mathbf{k},t)} = \left[n_a \chi_{aa} \tau_\text{hold} \cos(\theta) \sinc\left(\omega_\mathbf{k} \tau_\text{hold}\right) \right]^2, \label{eqBogOccupation}
\end{align}
where the Bogoliubov mode frequency is $\omega_\mathbf{k} = \sqrt{\omega^0_\mathbf{k}(\omega^0_\mathbf{k} + 2 \chi_{aa} n_a)}$, and the free particle frequency is $\omega^0_\mathbf{k} = \hbar \mathbf{k}^2/2 M$.  The Bogoliubov modes can be expected to have an impact on the squeezing of the system when their occupation is a non-negligible fraction of the total number of atoms.  The fractional occupation of the Bogoliubov modes is
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &= \frac{1}{N}\sum_{\mathbf{k}\neq 0} n_a(\mathbf{k}, t), \label{eqCondensateDepletion}
\end{align}
where the sum over $\mathbf{k} \neq 0$ is taken over the available non-zero momentum modes $\mathbf{k} = (2\pi/L_x) n_x \hat{\mathbf{x}} + (2\pi/L_y) n_y \hat{\mathbf{y}} + (2\pi/L_z) n_z \hat{\mathbf{z}}$ with $n_x$, $n_y$ and $n_z$ arbitrary integers that are not all zero.

Figure~\ref{figBogSqueezingValidation} demonstrates the agreement between truncated Wigner simulations and the Bogoliubov theory for the fractional occupation of the Bogoliubov modes for parameters where the occupation of these modes approach 20\%\footnote{FIXME} of the total occupation of the system.  The disagreement for larger fractional occupations occurs because the Bogoliubov theory used in deriving \eqref{eqCondensateDepletion} is not number-conserving as it neglects the effect of the Bogoliubov modes on the condensate.

To minimise the damaging effects of the non-zero momentum modes it is better to operate in a regime where the total occupation of the non-zero momentum modes $N_{\mathbf{k}\neq 0}$ is small, which is also fortuitously the regime in which the Bogoliubov theory is a good approximation.  To better understand when $N_{\mathbf{k}\neq 0}/N$ can be expected to be small, we approximate the sum in \eqref{eqCondensateDepletion} with an integral.  In 1D Eq.~\eqref{eqCondensateDepletion} is
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &= \frac{(n_a \chi_{aa} \tau_\text{hold} \cos\theta)^2}{N} \sum_{k \neq 0} \sinc^2\left(\sqrt{\omega^0_{k}(\omega^0_{k}+2 \chi_{aa}n_a)}\tau_\text{hold}\right).
\end{align}
We wish to approximate this sum by the integral
\begin{align}
  \frac{N_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{(n_a \chi_{aa} \tau_\text{hold} \cos\theta)^2}{N} \int_{-\infty}^{+\infty} \sinc^2 \left(\sqrt{\omega^0_{k}(\omega^0_{k}+2 \chi_{aa}n_a)}\tau_\text{hold}\right) \frac{d k}{\Delta k}, \label{eqCondensateDepletionFirstIntegralApproximation}
\end{align}
where $\Delta k = 2\pi/L$.  This approximation will be valid in the limit that $\omega_{k_\text{min}} \tau_\text{hold} \ll 1$, where $k_\text{min} = 2\pi/L$ as for small $k$ the allowed values $k = n_x 2\pi /L$ will closely sample the region $\xi = \sqrt{\omega_k^0(\omega_k^0 + 2 \chi_{aa} n_a)} \tau_\text{hold} < \pi$ where $\sinc^2(\xi)$ is significant.

Changing integration variables to $\Phi = \omega^{0}_{\mathbf{k}}\tau_\text{hold}$ and defining $\Lambda_a = n_a \chi_{aa} \tau_\text{hold}$ the expression \eqref{eqCondensateDepletionFirstIntegralApproximation} can be written as
\begin{align}
  \frac{N^\text{1D}_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{1}{N} \frac{\Lambda_a^2 L \cos^2\theta } {\sqrt{2}\pi\sqrt{\frac{\hbar}{M}\tau_\text{hold}}} f^\text{1D}(\Lambda_a),
\end{align}
where
\begin{align}
  f^\text{1D}(\Lambda) &= \int_0^{\infty} \frac{1}{\sqrt{\Phi}}\sinc^2\sqrt{\Phi(\Phi + 2 \Lambda)}\, d\Phi.
\end{align}
In 2D and 3D the equivalent expressions are
\begin{align}
  \frac{N^{\text{2D}}_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{1}{N} \frac{\Lambda_a^2 A \cos^2\theta } {4\frac{\hbar}{M}\tau_\text{hold}} f^\text{2D}(\Lambda_a), \\
  \frac{N^{\text{3D}}_{\mathbf{k}\neq 0}(t)}{N} &\approx \frac{1}{N} \frac{\Lambda_a^2 V \cos^2\theta } {\sqrt{2}\pi^2\left(\frac{\hbar}{M}\tau_\text{hold}\right)^{3/2}} f^\text{3D}(\Lambda_a),
\end{align}
where
\begin{align}
  f^\text{2D}(\Lambda) &= \int_0^{\infty} \sinc^2\sqrt{\Phi(\Phi + 2 \Lambda)}\, d\Phi,\\
  f^\text{3D}(\Lambda) &= \int_0^{\infty} \sqrt{\Phi} \sinc^2\sqrt{\Phi(\Phi + 2 \Lambda)}\, d\Phi.
\end{align}

@@

Qualitatively, the occupation of the non-zero momentum modes is almost entirely determined by the number of Bogoliubov modes energetically accessible to the system.  While this gets larger

The Bogoliubov theory used in deriving \eqref{eqCondensateDepletion} is not number-conserving

To minimise the damaging effects of the non-zero momentum modes, it is better to

The Bogoliubov theory used in deriving \eqref{eqBogOccupation} is not number-conserving because the effect of the Bogoliubov modes on the condensate is neglected.

To minimise the

Good squeezing will only be observed when the fractional occupation of the Bogoliubov modes $N_{\mathbf{k}\neq 0}/N$ is small.

% I'm commenting this out and will essentially gloss over the matter.  We will illustrate that Bog modes are detrimental, rather than arguing the fact.
% [Insert motivation that the impact of Bogoliubov modes is detrimental here.  I could wave my hands and say that more modes is bad, as seen in the previous parts of the article, or I could refer to the squeezing of the Bogoliubov modes:
% \begin{align}
%   S_{a,\mathbf{k}\neq 0}(t_3) &= \frac{\frac{1}{2}\sum_{\mathbf{k}\neq0}n_a(\mathbf{k},t_3)\left[3 + \cos 2\theta + n_a(\mathbf{k}, t_3)\right]}{\sum_{\mathbf{k}\neq0} n_a(\mathbf{k}, t_3)},
% \end{align}
% which is always greater than 1, but this isn't convincing either as the total squeezing includes the number variance of the $\mathbf{k=0}$ mode as well as the correlations between the $\mathbf{k}=0$ mode and the $\mathbf{k}\neq 0$ modes.  If you do it naively, you find that when the Bog modes have a significant impact on the squeezing, they have also significantly changed it at $t=t_2$, which must be the same as $t=t_1$.  The problem is that the Bogoliubov theory used isn't number-conserving.  The best argument I have at the moment for Bog theory being a good indicator for when things go bad is that a significant Bog fraction implies a depleted condensate, and that will almost certainly make squeezing bad as it will invalidate the single-mode theory.  More freedom == bad?  In this case, the best evidence for the utility of Bog theory would seem to be Figure~\ref{fig1D_squeezing_and_bog_scaling}.]

@@

\begin{figure}
    \centering
    \framebox{
    \parbox{0.8\textwidth}{
    Plot of the number occupation of the condensate and non-zero momentum modes as computed by the Bog theory (direct summation and integral approximation) and a simulation.  The simulation agrees with the Bog theory for the occupation of the non-zero momentum modes, but there is significant ~10-20\% depletion of the condensate.  The horizontal axis is the hold time.}
    }
\caption{This is not a caption.}
  \label{figBogSqueezingValidation}
\end{figure}
